\large How resource abundance and stochasticity affect animals' spatial needs

### 2022-11-14 writing/manuscript.Rmd
resolved conflicts

<!--
TO DO:
- write conclusion paragraph
- write discussion (2-3 pages)
  - add headers for sections
- focus on V(R) as temporal variance rather than spatial one. people have already focused on patchiness, but temporal variation is more risky
-->

<!-- add custom commands for formulae -->

\newcommand{\e}{\mathbb E}
\newcommand{\var}{\mathbb V}
\newcommand{\vpos}{\mathbb V(\text{position})}

<!-- have table of contents on pages of its own -->

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = '75%', fig.align = 'center')
```

@@

<!--
@ciuti and @berger-tal: human presence in nature for other reasons than hunting
"When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
habitat productivity hypothesis: Harestad and Bunnel, 1979 [@harestad_home_1979]
-->

@@

<!-- You need a lot more depth here. Open with a paragraph on movement generally that leads to the idea that responses to resource are something worth focusing on. Follow that up with a paragraph on the history of exploring movement/space use and resources, leading to the concept that, to date, the relationship between abundance and stochasticity have not been fully explored. Then finish with a paragraph detailing the aims of the current paper. -->

\noindent The amount of resources an organism is able to access is a strong determinant of its probability to survive and reproduce. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], or even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms will change their behaviors in response to changes in local resource abundance. Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].

While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce. Two years later, @harestad_home_1979 proposed that the simplest hypothesis for the relationship between an animal's home range size is inversely proportional to resource abundance. More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022]. Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020]. Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists, including @lindstedt_home_1986 and @reiss_scaling_1988. Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.

@@

The hypothesis that spatial needs decrease with resource abundance is common and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @bista_effect_2022]. This is problematic because it can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between spatial needs and resource abundance must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime and the minimum amount of space it requires to survive (which is necessarily greater than zero). The relationship between resource abundance and animals' spatial needs is explored in more depth in a following section.

@@

where $\nu$ indicates the NDVI value. Note that only **linear transformations** (i.e. only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution's parameters [@denny_fallacy_2017; @jensen_sur_1906]. In this case, we can define $R$ as following a Beta distribution using the notation $B (\mu(t), \sigma^2(t))$. Failing to model or simulate resource abundance appropriately can fail to produce robust, sensible, and accurate results.

@@

such that an animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec u_t$), as well as how each of the parameters may change with the current environmental factors ($\vec r_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec r_t, \text{and }\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels), doing so with sufficient frequency and accuracy would not be simple. However, we can focus on animals' behavior and spatial needs as a response to resource abundance and unpredictability alone (rather than measuring changes in each of an animal's resources, needs, and behavioral states directly) <!-- since accounting for a proxy of resource abundance will also account for the effects of many other drivers that depend strongly on resource abundance (e.g., competition, predation)-->.

@@

# Effects of resource abundance on animals' spatial needs

```{r mu-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\var(\\text{position})$) of an animal as a function of mean resource abundance ($\\e(R)$). We expect low values of $\\e(R)$ to result in high $\\var(\\text{position})$ as animals are forced to explore greater areas to collect the resources they require to survive, whether they be range-resident or nomadic. As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive. Note that the relationship between $\\e(R)$ and $\\var(\\text{position})$ cannot be linear because it would require $\\var(\\text{position})$ to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

@@

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@baldwin_nutritional_1984; @ref]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [see see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @ex]. Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

@@

# Effects of environmental variance on animals' spatial needs

```{r s2-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\var(\\text{position})$) of an animal as a function of resource unpredictability ($\\var(R)$). We expect low values of $\\var(R)$ to result in low $\\var(\\text{position})$ as animals are able to depend on somewhat predictable resources. As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration. Note that the relationship between $\\var(R)$ and $\\var(\\text{position})$ cannot be linear because it would require $\\var(\\text{position})$ to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
```

Although it generally acknowledged that animals will modify their behavior in response to human activity [@rogala_human_2011] as well as changing conditions [@walther_ecological_2002; @chevin_adaptation_2010; @sih_evolution_2011], a fine-scale understanding of behavioral responses is lacking [@sih_evolution_2011; @beever_behavioral_2017].

@@

@@

This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case [@refs; but see @refs]. Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response. Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' $\vpos$ should be estimated before assuming the effect is negligible.

@@

*Needs for understanding effects of variance* are compounded by climate change, which exposes species to increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018]. Furthermore, anthropogenic structures reduce the habitat available to terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].

@@

\noindent Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$. The top row (constant $\var(R)$) shows how  remains constant. Some noise in $H$ is still present since $\var(R) \ne 0$ and thus $R$ is not constant). Note the nonlinear decrease in $H$ when $\e(R)$ increases at a constant slope (linear $\e(R)$<!--, constant $\var(R)$-->), with larger changes when $\e(R)$ is low, since the multiplicative changes in $R$ are larger. Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$. In regions where the average resource abundance changes over time (cyclical $\e(R)$<!--, constant $\var(R)$-->), animals should have sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$. However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unpredictable changes in $\e(R)$ may increase animals' spatial needs unpredictably. Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the expected spatial needs. This is particularly the case in environments where resource abundance changes unpredictably (drifting $\e(R)$<!--, constant $\var(R)$-->), since accurate long-term estimates of $\e(R)$ may be hard to produce, if not impossible. In cases where $\e(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly and often. Finally, in ecosystems where $\e(R)$ changes erratically and suddenly (erratic $\e(R)$<!--, constant $\var(R)$-->), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.

```{r 5-5-sims, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource availability ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time. The bottom line indicates the animal's core home range (0.5 quantile), while the top line indicates the 0.95 utilization quantile. Note how both quantiles decrease nonlinearly as $\\e(R)$ increases, and they increase approximately linearly as $\\var(R)$ increases. Additionally, the variance in both quantiles is higher when $\\var(R)$ is higher, and changes in $\\var(R)$ have greater impacts when $\\e(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the point $\\langle 0, 0\\rangle$ until they reach satiety, at which point they returned to $\\langle 0, 0\\rangle$ over the same amount of time. The animal's spatial variance parameter was then calculated using an OUF model via the $\\tt{ctmm}$ package", out.width='100%'}
knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
```

The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(H)$) illustrates how an animal's home range may vary as $\var(R)$ changes over time while $\e(R)$ remains constant. Both $\e(H)$ and $\var(H)$ increase with $\var(R)$, although $H_{95\%}$ is more sensitive to changes in $\var(R)$ than $H_{50\%}$. Additionally, the effect of $\var(R)$ on $H$ depends on $\e(R)$, since the effect of $\var(R)$ on $H$ is stronger when $\e(R)$ is low (see the bottom row with erratic $\var(R)$).

Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios. $\e(R)$ and $\var(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\e(R)$ and $\var(R)$ remain undetected. Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly continuously, such increases may be possible for short periods of time (followed by periods of no change or decrease). Additionally, these examples are important because they demonstrate the relationships between $H$, $\e(R)$, and $\var(R)$ in (relatively) simple scenarios. Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity. Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation). However, if highly stochastic or erratic changes occur frequently, animals may perceive them as smooth changes in $\e(R)$ or increases in $\var(R)$ rather a series of small, sudden, changes. Estimating the true trend in $\e(R)$ and $\var(R)$ would often require an excessively high cognitive capacity and an equally unlikely abundance of information, since although changes in $\e(R)$ are not due to $\var(R)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020], and $\e(R)$ and $\var(R)$ are not independent for $R \sim \Gamma(\mu, \sigma^2)$. <!-- https://doi.org/10.1080/00207549508930162 ?-->

[*Needs for understanding effects of variance*] are compounded by climate change, which exposes species to increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018]. Furthermore, anthropogenic structures reduce the habitat available to terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].

@@

## Moving window examples

@@

>>>>

<!-- define custom math symbols -->

\newcommand*\e{\mathbb E} <!-- expectation (blackboard bold) E -->

\newcommand*\var{\mathbb V} <!-- variance (blackboard bold) V -->

\newcommand*\vpos{\var (\text{position})} <!-- positional variance -->

<!-- place table of contents on a new page -->
\newpage

@@

```{r setup, include=FALSE}
# set chunk defaults: don't print code, images 75% of page width and centered
knitr::opts_chunk$set(echo = FALSE, out.width = '75%', fig.align = 'center')
```

@@

<!--
@ciuti and @berger-tal: human presence in nature for reasons other than hunting
"When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
habitat productivity hypothesis: Harestad and Bunnel, 1979 [@harestad_home_1979]
-->

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its probability to survive and reproduce. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], or even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms will change their behaviors in response to changes in local resource abundance. Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].

While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce. Two years later, @harestad_home_1979 proposed that the simplest hypothesis for the relationship between an animal's home range size is inversely proportional to resource abundance. More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022]. Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020]. Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists, including @lindstedt_home_1986 and @reiss_scaling_1988. Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs. **but animals don't only respond to changes in mean**

**paragraph detailing effects of variance on movement and how little work has been done**

@@

The hypothesis that spatial needs decrease with resource abundance is common and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @bista_effect_2022; @bradsworth_using_2022]. This is problematic because it can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between spatial needs and resource abundance must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime and the minimum amount of space it requires to survive (which is necessarily greater than zero). The relationship between resource abundance and animals' spatial needs is explored in more depth in a following section.

@@

where $\nu$ indicates the NDVI value. Note that only **linear transformations** (i.e. only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution's parameters [@denny_fallacy_2017; @jensen_sur_1906]. In this case, we can define $R$ as following a Beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$ (We use this parameterization here for ease of explanation, but note that Beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is $\mu = \frac{\alpha}{\alpha+\beta}$ while the variance is $\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$). Failing to model or simulate resource abundance appropriately can fail to produce robust, sensible, and accurate results.

@@

such that an animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec u_t$), as well as how each of the parameters may change with the current environmental factors ($\vec r_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec r_t,$ and $\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels), doing so with sufficient frequency and accuracy would not be simple. However, we can focus on animals' behavior and spatial needs as a response to resource abundance and unpredictability alone (rather than measuring changes in each of an animal's resources, needs, and behavioral states directly).

@@

# Effects of $\e(R)$

```{r mu-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\vpos$) of an animal as a function of mean resource abundance ($\\e(R)$). We expect low values of $\\e(R)$ to result in high $\\vpos$ as animals are forced to explore greater areas to collect the resources they require to survive, whether they be range-resident or nomadic. As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive. Note that the relationship between $\\e(R)$ and $\\vpos$ cannot be linear because it would require $\\vpos$ to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

@@

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@baldwin_nutritional_1984; @reich_body_2001; @ref]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @ex]. Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

@@

# Effects of $\var(R)$

```{r s2-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\vpos$) of an animal as a function of resource unpredictability ($\\var(R)$). We expect low values of $\\var(R)$ to result in low $\\vpos$ as animals are able to depend on somewhat predictable resources. As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration. Note that the relationship between $\\var(R)$ and $\\vpos$ cannot be linear because it would require $\\vpos$ to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
```

@@

# Interactive effects of $\e(R)$ and $\var(R)$

@@

This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance. Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case [@refs; but see @refs]. Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response. Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' $\vpos$ should be estimated before assuming the effect is negligible. The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018]. Additionally, anthropogenic structures and urban areas reduce the habitat available to many terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018]. Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable [@ref] and pose high mortality risks [@noonan_roads_2021; @refs].

@@

Most habitats with seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ often changes both within and between years. In such cases, failing to account for $\var(R)$ can result in serious biases in the estimation of animals' spatial needs [@examples]. In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or in periods of increased energetic needs, such as reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [@male_deer_growing_antlers]. In heterogeneous environments, such as habitats with a high diversity of clustered resources, animals may have to move between locations if they cannot depend on a single one.

@@

\noindent Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$. The top row (constant $\var(R)$) shows how $\vpos$ remains constant. Some noise in $H$ is still present since $\var(R) \ne 0$ and thus $R$ is not constant). Note the nonlinear decrease in $H$ when $\e(R)$ increases at a constant slope (linear $\e(R)$), with larger changes when $\e(R)$ is low. This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g. from 1 to 2) can correspond to large multiplicative changes (i.e., 2 is twice as big as 1). Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$ since larger quantiles (e.g., 95% > 50%) are more sensitive to changes in the $\vpos$. In regions where the average resource abundance changes over time (cyclical $\e(R)$; e.g., seasonal changes), animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$ (e.g., winter park closures). However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unpredictable changes in $\e(R)$ may increase animals' spatial needs unpredictably. Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the estimated spatial needs. This is particularly the case in environments where resource abundance changes unpredictably (drifting $\e(R)$), since accurate long-term estimates of $\e(R)$ may be hard to produce, if not impossible. In cases where $\e(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time. Finally, in ecosystems where $\e(R)$ changes erratically and suddenly (erratic $\e(R)$), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.

```{r 5-5-sims, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource availability ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time. The bottom line indicates the animal's core home range (0.5 quantile), while the top line indicates the 0.95 utilization quantile. Note how both quantiles decrease nonlinearly as $\\e(R)$ increases, and they increase approximately linearly as $\\var(R)$ increases. Additionally, the variance in both quantiles is higher when $\\var(R)$ is higher, and changes in $\\var(R)$ have greater impacts when $\\e(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the point $\\langle 0, 0\\rangle$ until they reach satiety, at which point they returned to $\\langle 0, 0\\rangle$ over the same amount of time. The animal's spatial variance parameter was then calculated using an Ornstein-Uhlenbeck Foraging (OUF) model via the $\\tt{ctmm}$ package", out.width='100%'}
knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
```

Both $\e(H)$ and $\var(H)$ increase with $\var(R)$, although the 95% home range ($H_{95\%}$) is more sensitive to changes in $\var(R)$ than the core home range ($H_{50\%}$). Additionally, the effect of $\var(R)$ on $H$ depends on $\e(R)$, since the effect of $\var(R)$ on $H$ is stronger when $\e(R)$ is low (see the bottom row with erratic $\var(R)$).

Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios. $\e(R)$ and $\var(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\e(R)$ and $\var(R)$ remain undetected. Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly continuously, such increases may be possible for short periods of time (followed by periods of no change or decrease). Additionally, these examples are important because they demonstrate the relationships between $H$, $\e(R)$, and $\var(R)$ in (relatively) simple scenarios. Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity. Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation). However, if highly stochastic or erratic changes occur frequently, animals may perceive them as smooth changes in $\e(R)$ or increases in $\var(R)$ rather a series of small, sudden, changes. Estimating the true trend in $\e(R)$ and $\var(R)$ would often require an excessively high cognitive capacity and an equally unlikely abundance of information, since although changes in $\e(R)$ are not due to $\var(R)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020], and $\e(R)$ and $\var(R)$ are not independent for $R \sim \Gamma(\mu, \sigma^2)$.

```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements. The relationships were estimated using a Generalized Linear Model with a Gamma family. The thicker bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the thinner top line indicates the relationship with the 0.95 utilization quantile. Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlineear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
knitr::include_graphics('../figures/simulation-regression-plots.png')
```

@@

## Applying this framework

* The hypotheses we present here allow researchers to combine findings and generalize
* the analytical methods we use provide an example for how researchers can test these hypotheses using their own data
* can present results building on previous work rather than listing them as *de novo* findings
* additionally, can test for common behaviors between members of the same population or species
* ecologists often conclude that populations from different regions behave differently [@examples], but their differences in behavior may only be due to differences in $\e(r)$ or $\var(R)$
* using the methods presented here, it is possible to test whether populations and individuals do behave differently while also quantifying the differences in behavior

### Modeling $R$

* location-scale models
* `mgcv`, `brms`
* choosing the best family

### Modeling spatial needs

* importance of AKDE
* Introduce moving window idea

Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; bald eagles: @wheat_migrate_2017]. However, the models we use in this paper assume an animal's large-scale behavior and movement to remain consistent over time...

### Modeling the relationship

* Gamma distribution for HR
* possibly location-scale if enough data?

## *Tapirus terrestris* in the Brazilian Pantanal: an empirical example

@@

# MISC

In the case of animals whose spatial needs change over the their lifetimes [@refs], one may define HR as a function of time with distinct HRs for each period [e.g., salmon spawning: @ref] or as a continuous function of age [so the HR changes smoothly: @ref]. In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019]. Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available.
